Small Language Modelling to Generate Playlists from Song Sequences

### 2023-12-12 docs/draft_report.tex
Final Draft

\author{
  Mark Green\\
  Data Science Program\\
  Indiana University\\
  Bloomington, IN 47405\\
  \texttt{margree@iu.edu}\\
  \And
  Sahil Dhingra \\
  Data Science Program\\
  Indiana University\\
  Bloomington, IN 47405\\
  \texttt{sahdhin@iu.edu}\\
  \And
  Rahul Jain \\
  Data Science Program\\
  Indiana University\\
  Bloomington, IN 47405\\
  \texttt{rjdharmc@iu.edu}\\
}

@@

\begin{abstract}
  Large Language Models (LLMs) such as GPT and BERT have led to revolutionary advances in language applications such as machine translation, sentiment analysis, and text generation. However, these models rely on enormous datasets of texts derived from spoken language which are prohibitively resource-intensive to train. Although pre-trained models solve this to a degree, their scope is limited to the pre-trained spoken language so they cannot be applied to more niche language modeling tasks. This paper explores the application of neural networks for small language modelling where the dataset is small in both size and vocabulary and is composed of non-spoken language “words”. Specifically, this approach explores the music of the band Phish, leveraging historical song appearances and setlists from their shows to model the band's performance tendencies. By decoding intricate patterns from their extensive performance history using a targeted small language model, this approach offers a unique solution for personalized music recommendations. The research aims to enhance the user experience for dedicated fans with more artist-centric music recommendation systems and to test the performance of neural network language models for small, niche applications.
\end{abstract}

@@

We purport that constructing small language models from the song sequences of live performances to generate playlists for a given band could increase the engagement of “dedicated fan” type users by generating playlists that more closely mimic a band’s musical tendencies than a more generalized algorithm. In this report, we explore the music of the Jam band Phish for this purpose – who has not only a large catalog of live shows and songs, but also a very dedicated fanbase. A previous endeavor\footnote{\url{https://towardsdatascience.com/predicting-what-song-phish-will-play-next-with-deep-learning-947ccce3824d}} employed a Long Short-Term Memory (LSTM) model with Word2Vec [4] algorithm for token embeddings to achieve a approxmately 21\% accuracy for next song prediction given a sequence of Phish songs. This paper builds upon this work by recreating it four years later and exploring the prediction task with alternate embedding methods, and RNN model architectures, and attention model architectures.

@@

The first stage of experimentation was an attempt to reproduce the original experiment\footnote{\url{https://github.com/andrewrreed/phish-setlist-modeling}} from November 2019 with the latest setlist data from December 2023. Experiments were run for multiple model architectures, both with and without Word2Vec embedding. For both architectures, a grid search was performed with song sequence lengths of 50 and 150 with varying LSTM unit sizes and dropout rates. Although the original experiment reported the best next song accuracy of approximately 21.8\%, We were unable to reproduce these results and none of the 16 models attempted in the grid search produced satisfactory results with Dec 2023 dataset. The model without Word2Vec embedding suffered from an exploding gradient problem, and the validation accuracy quickly approached 0 after just a few training epochs. Models with Word2Vec embedding did marginally better, but would also become saturated at low validation accuracy within 4-5 training epochs. These results indicate the model used in the original experiment was overfitting the data from Nov 2019, and hence did not generalize well to the latest data. The highest validation accuracies observed were approximately 12\% for models without Word2Vec and 10\% for the models with Word2Vec embedding - both significantly less than the originally reported accuracy of 21.8\%. This highlights the difficulty of using only the unidirectional songs sequences to predict new song sequences, and  suggests more context and data is needed to obtain a more meaningful inference. This conclusion is in line with the original Author's assessment as well, and we attempt to address this in further experiments.

@@

In the initial exploratory data analysis of the Phish concert dataset, it was revealed the frequency of songs played resembled a power-law distribution whereby almost half of the songs were only played once or twice throughout the band's entire history. To address the challenge posed by rarely performed songs, we consolidated these one-off and two-off songs into a unified song category termed the "wildcard". For prediction, this wildcard song is treated like any other song Phish plays but it represents an intractable uncertainty to relax the problem for new song debuts or unpredictably random cover songs.

@@

Recall, the aim of this project is to generate song sequences (IE playlists) which are representative of the style in which the band plays. This problem can be framed in terms of language modelling in either a {\it sequence-to-token} perspective, using a sequence to predict the next token, or a {\it sequence-to-sequence} perspective, using a sequence to predict the next sequence. Since the models will have to either predict the next $n^{th}$ song or the next $n$ songs, multiple data splitting techniques for input and output data were applied to evaluate the model performance under both of these paradigms. These splitting mechanisms are described in the following section, specific for each model.

@@

Since the language modelling is trying to find good features rooted in the sequential aspect of the data, both RNN LSTM and attention architectures were modeled so as to focus on the sequential information to make better predictions. For RNN experiments, we used both uni-directional and bi-directional models were attempted. For the attention experiments, self-attention and cross-attention were attempted. These model architectures are discussed in more detail below.

@@

\paragraph{Sequential Training Data}
The training data was structured sequentially as [n], [n, n+1], [n, n+1, n+2], and so on, with corresponding labels as [n+1], [n+2], [n+3], and so forth. Emphasizing continuity, this method enables the model to learn and predict the sequential progression of song sequences.

@@

In contrast to the sequence-to-token models attempted with the LSTM architectures, sequence-to-sequence models are attempted with multi-head attention architectures. These models process the entire song sequence for a setlist into offset pairs to form the inputs and target outputs, where the original sequence is broken down into [0, ..., $n-1$] tokens for the input sequence and [1, ..., $n$] tokens for the output sequence. The sequences can be any size, although sequence lengths of 125 songs performed the best. The token sequences are encoded by an embedding layer composed of the sum of a learned N-dimensional vector embedding layer and a positional embedding layer. The positional embedding layer applies sine and cosine functions for fixed positional encodings. This is the same mechanism as is used in the original attention model [5].

The model architecture itself is similar to the decoder-only transformer used in the GPT model[6, 7]. It consists of a causal self-attention layer using masked multi-head attention to attend to the sequencing. This layer is added and normalized, then passed to a feed forward layer to form the Transformer block. These transformer blocks are then layered sequentially and an additional feed forward layer computes the final output, then activated by softmax.

@@

This final model architecture is setup similarly to the original transformer model proposed in "Attention is All You Need” [5]. Here, a sequence of songs is concatenated from the previous 5 or more setlists to be used as the context for the global self-attention layer, which is passed to the decoder via the cross-attention layer. Inputs to be predicted upon are first sent through the causal self-attention layer as part of the decoder, before then being sent through the cross-attention layer for the given context sequence to form the transformer block. These transformer blocks are similarly layered and fed forward to compute the final output, activated by softmax to generate the multi-class probability classification.

@@

\begin{table}
	\caption{Summary of Model Training Results}
	\label{results-table}
	\centering
	\begin{tabular}{lll}
		\toprule
		Model & Best Next-Song Validation Accuracy & Generated Sequence Quality \\
		\midrule
		2019 Model Recreation & $\sim$12.5\% & NA\footnote{Setlist generation was not tested for the 2019 model recreation attempts.} \\
		Model 1 & $\sim$18\% & Poor \\
		Model 2 & 19.5\% & Moderate \\
		Model 3 & 19.0\% & Moderate \\
		Model 4 & 16.8\% & Good \\
		\bottomrule
	\end{tabular}
\end{table}

@@

For each model, the setlist used for comparison was:

@@

The question is how good is this accuracy? Considering the probability of random chance, and problem set, the Bayesian probability greater than 11\% is an improvement. we expected to get further improvement on the accuracy but we didn't cross the 24\% mark. Further experiment of feeding other features as categorical data to model could possibly be a way forward, which might provide a little better insight into the playlist. In addition, having more data on genreof music, , instruments etc could have helped shape the playlist better.

These experiments were unable to reproduce the original experiment and were also unable to achieve resulting next-word accuracies exceeding those of the original experiment. However, there was some moderate success with generating song sequences in ways that could have been created by the band. In this way, certain structures of the song sequences were learned.

@@

Some concluding words.

@@

	[5] A. Vaswani, N. Shazeer, N. Parmar, J. Uszkoreit, L. Jones, A. N. Gomez, L. Kaiser, I. Polosukhin, “Attention Is All You Need.” arXiv, 2017. doi: 10.48550/ARXIV.1706.03762.

	[6] A. Radford, J. Wu, R. Child, D. Luan, D. Amodei, I. Sutskever. "Language models are unsupervised multitask learners". 2019.

	[7] A. Radford, K. Narasimhan, T. Salimans, I. Sutskever. "Improving Language Understanding by Generative Pre-Training". 2017.
}

\section{Submission of papers to NeurIPS 2023}

Please read the instructions below carefully and follow them faithfully. \textbf{Important:} This year the checklist will be submitted separately from the main paper in OpenReview, please review it well ahead of the submission deadline: \url{https://neurips.cc/public/guides/PaperChecklist}.

\subsection{Style}

Papers to be submitted to NeurIPS 2023 must be prepared according to the
instructions presented here. Papers may only be up to {\bf nine} pages long,
including figures. Additional pages \emph{containing only acknowledgments and
references} are allowed. Papers that exceed the page limit will not be
reviewed, or in any other way considered for presentation at the conference.

The margins in 2023 are the same as those in previous years.

Authors are required to use the NeurIPS \LaTeX{} style files obtainable at the
NeurIPS website as indicated below. Please make sure you use the current files
and not previous versions. Tweaking the style files may be grounds for
rejection.

\subsection{Retrieval of style files}

The style files for NeurIPS and other conference information are available on
the website at
\begin{center}
  \url{http://www.neurips.cc/}
\end{center}
The file \verb+neurips_2023.pdf+ contains these instructions and illustrates the
various formatting requirements your NeurIPS paper must satisfy.

The only supported style file for NeurIPS 2023 is \verb+neurips_2023.sty+,
rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{} 2.09,
  Microsoft Word, and RTF are no longer supported!}

The \LaTeX{} style file contains three optional arguments: \verb+final+, which
creates a camera-ready copy, \verb+preprint+, which creates a preprint for
submission to, e.g., arXiv, and \verb+nonatbib+, which will not load the
\verb+natbib+ package for you in case of package clash.

\paragraph{Preprint option}
If you wish to post a preprint of your work online, e.g., on arXiv, using the
NeurIPS style, please use the \verb+preprint+ option. This will create a
nonanonymized version of your work with the text ``Preprint. Work in progress.''
in the footer. This version may be distributed as you see fit, as long as you do not say which conference it was submitted to. Please \textbf{do
  not} use the \verb+final+ option, which should \textbf{only} be used for
papers accepted to NeurIPS.

At submission time, please omit the \verb+final+ and \verb+preprint+
options. This will anonymize your submission and add line numbers to aid
review. Please do \emph{not} refer to these line numbers in your paper as they
will be removed during generation of camera-ready copies.

The file \verb+neurips_2023.tex+ may be used as a ``shell'' for writing your
paper. All you have to do is replace the author, title, abstract, and text of
the paper with your own.

The formatting instructions contained in these style files are summarized in
Sections \ref{gen_inst}, \ref{headings}, and \ref{others} below.

\section{General formatting instructions}
\label{gen_inst}

The text must be confined within a rectangle 5.5~inches (33~picas) wide and
9~inches (54~picas) long. The left margin is 1.5~inch (9~picas).  Use 10~point
type with a vertical spacing (leading) of 11~points.  Times New Roman is the
preferred typeface throughout, and will be selected for you by default.
Paragraphs are separated by \nicefrac{1}{2}~line space (5.5 points), with no
indentation.

The paper title should be 17~point, initial caps/lower case, bold, centered
between two horizontal rules. The top rule should be 4~points thick and the
bottom rule should be 1~point thick. Allow \nicefrac{1}{4}~inch space above and
below the title to rules. All pages should start at 1~inch (6~picas) from the
top of the page.

For the final version, authors' names are set in boldface, and each name is
centered above the corresponding address. The lead author's name is to be listed
first (left-most), and the co-authors' names (if different address) are set to
follow. If there is only one co-author, list both author and co-author side by
side.

Please pay special attention to the instructions in Section \ref{others}
regarding figures, tables, acknowledgments, and references.

\section{Headings: first level}
\label{headings}

All headings should be lower case (except for first word and proper nouns),
flush left, and bold.

First-level headings should be in 12-point type.

\subsection{Headings: second level}

Second-level headings should be in 10-point type.

\subsubsection{Headings: third level}

Third-level headings should be in 10-point type.

\paragraph{Paragraphs}

There is also a \verb+\paragraph+ command available, which sets the heading in
bold, flush left, and inline with the text, with the heading followed by 1\,em
of space.

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone.

\subsection{Citations within the text}

The \verb+natbib+ package will be loaded for you by default.  Citations may be
author/year or numeric, as long as you maintain internal consistency.  As to the
format of the references themselves, any style is acceptable as long as it is
used consistently.

The documentation for \verb+natbib+ may be found at
\begin{center}
  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
\end{center}
Of note is the command \verb+\citet+, which produces citations appropriate for
use in inline text.  For example,
\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}
produces
\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

If you wish to load the \verb+natbib+ package with options, you may add the
following before loading the \verb+neurips_2023+ package:
\begin{verbatim}
   \PassOptionsToPackage{options}{natbib}
\end{verbatim}

If \verb+natbib+ clashes with another package you load, you can add the optional
argument \verb+nonatbib+ when loading the style file:
\begin{verbatim}
   \usepackage[nonatbib]{neurips_2023}
\end{verbatim}

As submission is double blind, refer to your own published work in the third
person. That is, use ``In the previous work of Jones et al.\ [4],'' not ``In our
previous work [4].'' If you cite your other papers that are not widely available
(e.g., a journal paper under review), use anonymous author names in the
citation, e.g., an author of the form ``A.\ Anonymous'' and include a copy of the anonymized paper in the supplementary material.

\subsection{Footnotes}

Footnotes should be used sparingly.  If you do require a footnote, indicate
footnotes with a number\footnote{Sample of the first footnote.} in the
text. Place the footnotes at the bottom of the page on which they appear.
Precede the footnote with a horizontal rule of 2~inches (12~picas).

Note that footnotes are properly typeset \emph{after} punctuation
marks.\footnote{As in this example.}

\subsection{Figures}

\begin{figure}
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

All artwork must be neat, clean, and legible. Lines should be dark enough for
purposes of reproduction. The figure number and caption always appear after the
figure. Place one line space before the figure caption and one line space after
the figure. The figure caption should be lower case (except for first word and
proper nouns); figures are numbered consecutively.

You may use color figures.  However, it is best for the figure captions and the
paper body to be legible if the paper is printed in either black/white or in
color.

\subsection{Tables}

All tables must be centered, neat, clean and legible.  The table number and
title always appear before the table.  See Table~\ref{sample-table}.

Place one line space before the table title, one line space after the
table title, and one line space after the table. The table title must
be lower case (except for first word and proper nouns); tables are
numbered consecutively.

Note that publication-quality tables \emph{do not contain vertical rules.} We
strongly suggest the use of the \verb+booktabs+ package, which allows for
typesetting high-quality, professional tables:
\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}
This package was used to typeset Table~\ref{sample-table}.

\begin{table}
  \caption{Sample table title}
  \label{sample-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule(r){1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Math}
Note that display math in bare TeX commands will not create correct line numbers for submission. Please use LaTeX (or AMSTeX) commands for unnumbered display math. (You really shouldn't be using \$\$ anyway; see \url{https://tex.stackexchange.com/questions/503/why-is-preferable-to} and \url{https://tex.stackexchange.com/questions/40492/what-are-the-differences-between-align-equation-and-displaymath} for more information.)

\subsection{Final instructions}

Do not change any aspects of the formatting parameters in the style files.  In
particular, do not modify the width or length of the rectangle the text should
fit into, and do not change font sizes (except perhaps in the
\textbf{References} section; see below). Please note that pages should be
numbered.

\section{Preparing PDF files}

Please prepare submission files with paper size ``US Letter,'' and not, for
example, ``A4.''

Fonts were the main cause of problems in the past years. Your PDF file must only
contain Type 1 or Embedded TrueType fonts. Here are a few instructions to
achieve this.

\begin{itemize}

\item You should directly generate PDF files using \verb+pdflatex+.

\item You can check which fonts a PDF files uses.  In Acrobat Reader, select the
  menu Files$>$Document Properties$>$Fonts and select Show All Fonts. You can
  also use the program \verb+pdffonts+ which comes with \verb+xpdf+ and is
  available out-of-the-box on most Linux machines.

\item \verb+xfig+ "patterned" shapes are implemented with bitmap fonts.  Use
  "solid" shapes instead.

\item The \verb+\bbold+ package almost always uses bitmap fonts.  You should use
  the equivalent AMS Fonts:
\begin{verbatim}
   \usepackage{amsfonts}
\end{verbatim}
followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or \verb+\mathbb{C}+
for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You can also use the following
workaround for reals, natural and complex:
\begin{verbatim}
   \newcommand{\RR}{I\!\!R} %real numbers
   \newcommand{\Nat}{I\!\!N} %natural numbers
   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
\end{verbatim}
Note that \verb+amsfonts+ is automatically loaded by the \verb+amssymb+ package.

\end{itemize}

If your file contains type 3 fonts or non embedded TrueType fonts, we will ask
you to fix it.

\subsection{Margins in \LaTeX{}}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+ from the \verb+graphicx+ package. Always specify the
figure width as a multiple of the line width as in the example below:
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
See Section 4.4 in the graphics bundle documentation
(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})

A number of width problems arise when \LaTeX{} cannot properly hyphenate a
line. Please give LaTeX hyphenation hints using the \verb+\-+ command when
necessary.

\begin{ack}
Use unnumbered first level headings for the acknowledgments. All acknowledgments
go at the end of the paper before the list of references. Moreover, you are required to declare
funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2023/PaperInformation/FundingDisclosure}.

Do {\bf not} include this section in the anonymized submission, only in the final paper. You can use the \texttt{ack} environment provided in the style file to autmoatically hide this section in the anonymized submission.
\end{ack}

\section{Supplementary Material}

Authors may wish to optionally include extra information (complete proofs, additional experiments and plots) in the appendix. All such materials should be part of the supplemental material (submitted separately) and should NOT be included in the main submission.

References follow the acknowledgments in the camera-ready paper. Use unnumbered first-level heading for
the references. Any choice of citation style is acceptable as long as you are
consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
when listing the references.
Note that the Reference section does not count towards the page limit.
\medskip

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
(eds.), {\it Advances in Neural Information Processing Systems 7},
pp.\ 609--616. Cambridge, MA: MIT Press.

[2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
  Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
recall at excitatory recurrent synapses and cholinergic modulation in rat
hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\author{
  Mark Green\\
  Data Science Program\\
  Indiana University\\
  Bloomington, IN 47405\\
  \texttt{margree}\\
  \And
  Sahil Dhingra \\
  Data Science Program\\
  Indiana University\\
  Bloomington, IN 47405\\
  \texttt{sahdhin}\\
  \And
  Rahul Jain \\
  Data Science Program\\
  Indiana University\\
  Bloomington, IN 47405\\
  \texttt{rjdharmc}\\
}

@@

\begin{abstract}
  Large Language Models (LLMs) such as GPT and BERT have led to revolutionary advances in language applications such as machine translation, sentiment analysis, and text generation. However, these models rely on enormous datasets of texts derived from spoken language which are prohibitively resource-intensive to train. Although pre-trained models solve this to a degree, their scope is limited to the pre-trained spoken language so they cannot be applied to more niche language modeling tasks. This paper explores the application of neural networks for small language modelling where the dataset is small in both size and vocabulary and is composed of non-spoken language “words”. Specifically, this approach explores the music of the band Phish, leveraging historical song appearances and setlists from their shows to model the band's performance tendencies. By decoding intricate patterns from their extensive performance history using a targeted small language model, this approach offers a unique solution for personalized music recommendations. The research aims to enhance the user experience for dedicated fans with more artist-centric music recommendation systems and to test the performance of neural network language models for small and niche applications.
\end{abstract}

@@

We purport that constructing small language models from the song sequences of live performances to generate playlists for a given band could increase the engagement of “dedicated fan” type users by generating playlists that more closely mimic a band’s musical tendencies than a more generalized algorithm. In this report, we explore the music of the Jam band Phish for this purpose – who has not only a large catalog of live shows and songs, but also a very dedicated fanbase. A previous endeavor\footnote{\url{https://towardsdatascience.com/predicting-what-song-phish-will-play-next-with-deep-learning-947ccce3824d}} employed a Long Short-Term Memory (LSTM) model with Word2Vec [4] algorithm for token embeddings to achieve a approxmately 21\% accuracy for next song prediction given a sequence of Phish songs. This paper builds upon this work by recreating it four years later and exploring the prediction task with alternate embedding methods, and RNN model architectures, and Attention model architectures.

@@

The first stage of experimentation was an attempt to reproduce the original experiment\footnote{\url{https://github.com/andrewrreed/phish-setlist-modeling}} from November 2019 with the latest setlist data from December 2023. Experiments were run for multiple model architectures, both with and without Word2Vec embedding. For both architectures, a grid search was performed with song sequence lengths of 50 and 150 with varying LSTM unit sizes and dropout rates. Although the original experiment reported the best next song accuracy of approximately 21.8\%, We were unable to reproduce these results and none of the 16 models attempted in the grid search produced satisfactory results with Dec 2023 dataset. The model without Word2Vec embedding suffered from an exploding gradient problem, and the validation accuracy quickly approached 0 after just a few training epochs. Models with Word2Vec embedding did marginally better, but would also become saturated at low validation accuracy within 4-5 training epochs. These results indicate the model used in the original experiment was overfitting the data from Nov 2019, and hence did not generalize well to the latest data. The highest validation accuracies observed were approximately 12\% for models without Word2Vec embedding layer and 10\% for the models with Word2Vec embedding - both significantly less than the originally reported accuracy of 21.8\%.

This highlights the difficulty of using only the unidirectional songs sequences to predict new song sequences, and  suggests more context and data is needed to obtain a more meaningful inference. This conclusion is in line with the original Author's assessment as well, and we attempt to address this in further experiments.

@@

In the initial exploratory data analysis of the Phish concert dataset, it was revealed the frequency of songs played resembled a \emph{power-law distribution} whereby almost half of the songs were only played once or twice throughout the band's entire history. To address the challenge posed by rarely performed songs, we consolidated these one-off and two-off songs into a unified song category termed the "wildcard". For prediction, this wildcard song is treated like any other song Phish plays but it represents an intractable uncertainty to relax the problem for new song debuts or unpredictably random cover songs.

@@

Recall, the aim of this project is to generate song sequences (i.e. playlists) which are representative of the style in which the band plays. This problem can be framed in terms of language modelling in either a {\it sequence-to-token} perspective, using a sequence to predict the next token, or a {\it sequence-to-sequence} perspective, using a sequence to predict the next sequence. Since the models will have to either predict the next $n^{th}$ song or the next $n$ songs, multiple data splitting techniques for input and output data were applied to evaluate the model performance under both of these paradigms. These splitting mechanisms are described in the following section, specific for each model.

@@

Since the language modelling is trying to find good features rooted in the sequential aspect of the data, both RNN LSTM and Attention architectures were modeled so as to focus on the sequential information to make better predictions. For RNN experiments, we used both uni-directional and bi-directional models were attempted. For experiments with Attention, self-Attention and cross-Attention were attempted. These model architectures are discussed in more detail below.

@@

\paragraph{Sequential Training Data}
The training data was structured sequentially as [$n$], [$n$, $n+1$], [$n$, $n+1$, $n+2$], and so on, with corresponding labels as [$n+1$], [$n+2$], [$n+3$], and so forth. Emphasizing continuity, this method enables the model to learn and predict the sequential progression of song sequences.

@@

In contrast to the sequence-to-token models attempted with the LSTM architectures, sequence-to-sequence models are attempted with multi-head Attention architectures. These models process the entire song sequence for a setlist into offset pairs to form the inputs and target outputs, where the original sequence is broken down into [0, ..., $n-1$] tokens for the input sequence and [1, ..., $n$] tokens for the output sequence. The sequences can be any size, although sequence lengths of 125 songs performed the best. The token sequences are encoded by an embedding layer composed of the sum of a learned N-dimensional vector embedding layer and a positional embedding layer. The positional embedding layer applies sine and cosine functions for fixed positional encodings. This is the same mechanism as is used in the original Attention model [5].

The model architecture itself is similar to the decoder-only transformer used in the GPT model[6, 7]. It consists of a causal self-Attention layer using masked multi-head Attention to attend to the sequencing. This layer is added and normalized, then passed to a feed forward layer to form the Transformer block. These transformer blocks are then layered sequentially and an additional feed forward layer computes the final output, then activated by softmax.

@@

This final model architecture is setup similarly to the original transformer model proposed in "Attention is All You Need” [5]. Here, a sequence of songs is concatenated from the previous 5 or more setlists to be used as the context for the global self-Attention layer, which is passed to the decoder via the cross-Attention layer. Inputs to be predicted upon are first sent through the causal self-Attention layer as part of the decoder, before then being sent through the cross-Attention layer for the given context sequence to form the transformer block. These transformer blocks are similarly layered and fed forward to compute the final output, activated by softmax to generate the multi-class probability classification.

@@

\begin{table}
	\caption{Summary of Model Training Results}
	\label{results-table}
	\centering
	\begin{tabular}{lll}
		\toprule
		Model & Best Next-Song Validation Accuracy & Generated Sequence Quality \\
		\midrule
		2019 Model Recreation & $\sim$12.5\% & Not Attempted \\
		Model 1 & $\sim$18\% & Poor \\
		Model 2 & 19.5\% & Moderate \\
		Model 3 & 19.0\% & Moderate \\
		Model 4 & 16.8\% & Good \\
		\bottomrule
	\end{tabular}
\end{table}

@@

For each model, the setlist used for comparison was as follows with the first three songs used as the seed:

@@

Considering the probability of random chance and given the small, diverse, and random nature of this "language" of song sequences, achieving above 16\% accuracy with multiple distinct model architectures shows these small language models are able to identify impactful features to a decent degree. Although we hoped to improve the accuracy beyond that obtained in the original 2019 experiment, we were unable to cross the 20\% next song accuracy mark. However, given we were unable to reproduce the original experiment to the same degree of accuracy as achieved previously, the results shown here significant. In particular, it is shown that Attention-based models are able to slightly outcompete the RNN LSTM model in terms of generative quality, and can be near peers in terms of next song accuracy.

Further experimentation on this subject could expand upon this work in the following ways:
\begin{itemize}
	\item Giving the models additional context for songs, such as key, tempo, or lyric sentiment, could greatly increase the within setlist predictive capability by generating better song transitions.
	\item Giving the models additional context for shows, such as date, venue, or location, could also help to improve long-term temporal predictions. For example, 'Auld-Lang-Syne' is always and only ever performed on new year's eve. 
	\item Additional model architectures could be tried - in particular generative adversarial networks (GANs) and conditioned variational autoencoders (CVAE) models would likely excel in the generation part.
	\item Dynamic Temporal Graphs could be used to model changing probabilities over time, and modelling song sequences as random walks with the prior temporal information could be a promising avenue to generate new setlists and predict song sequences [8, 9]. 
	\item Audio signal sequencing to learn sentiments and transitional markers for songs could be another way to improve song to song prediction accuracy. 
	\item This approach could be extended to other artist's. Although Phish has a large catalog of music, The Grateful Dead in particular has a similarly sized catalog of music and tendencies which could be interesting to explore. 
\end{itemize}

@@

The greatest obstacle for small language models is the size of the training dataset and the underlying structure of the language phenomenon which is being modeled. Despite this, RNNs and Attention-based transformer models can learn the syntactical tendencies of small languages to a limited degree. Despite being unable to recreate a 2019 project which attempted to model the music of Phish in this way, this project was able to reconstruct LSTM models to accomplish the same task with similar levels of accuracy and employ new Attention-based models to exceed the generative abilities of the RNN models. Additional work in this field should strongly consider the addition of more contextual information in the form of song or show attributes, temporal graph models, or audio signal sequencing.

@@

	[5] A. Vaswani, N. Shazeer, N. Parmar, J. Uszkoreit, L. Jones, A. N. Gomez, L. Kaiser, and I. Polosukhin, “Attention Is All You Need.” arXiv, 2017. doi: 10.48550/ARXIV.1706.03762.

	[6] A. Radford, J. Wu, R. Child, D. Luan, D. Amodei, and I. Sutskever. "Language models are unsupervised multitask learners". 2019.

	[7] A. Radford, K. Narasimhan, T. Salimans, and I. Sutskever. "Improving Language Understanding by Generative Pre-Training". 2017.

	[8] D. Gurevin, M. Shan, T. Geng, W. Jiang, C. Ding, and O. Khan, “Towards Real-Time Temporal Graph Learning.” arXiv, 2022. doi: 10.48550/ARXIV.2210.04114.

	[9] Y. Shen, J. Chen, P.-S. Huang, Y. Guo, and J. Gao, “M-Walk: Learning to Walk over Graphs using Monte Carlo Tree Search.” arXiv, 2018. doi: 10.48550/ARXIV.1802.04394.
}
